\textit{Associations between chlorophyll \emph{a} and various microcystin-LR health advisory concentrations}

### 2016-01-25 manuscript/f1000_submission/manuscript.Rmd
updated letter

---
title: Associations between chlorophyll *a* and various microcystin-LR health advisory
  concentrations
author:
- affilnum: 1
  email: hollister.jeff@epa.gov
  name: Jeffrey W. Hollister
- affilnum: 1
  name: Betty J. Kreakie
affiliation:
- affil: "US Environmental Protection Agency, Office of Research and Development, National
    Health and Environmental Effects Research Laboratory, Atlantic Ecology Division,
    27 Tarzwell Drive  Narragansett, RI, 02882, USA"
  affilnum: 1
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: f1000research.latex
  html_document: null
capsize: normalsize
csl: plos.csl
documentclass: article
fontsize: 11pt
linenumbers: yes
bibliography: manuscript.bib
spacing: doublespacing
abstract: "Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins. To help protect against these impacts, several health advisory levels have been set for some toxins. In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use. However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and advanced understanding required to interpret results. Addressing these issues will take time and resources. Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and in situ, and used as a first defense against high levels of microcystin. Chlorophyll *a* is commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin. In this paper, we use this association to provide estimates of chlorophyll a concentrations that are indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR. Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level. We look at the recent US EPA health advisories for drinking water as well as the World Health Organization levels for drinking water and recreational use and identify a range of chlorophyll *a* thresholds. A 50% chance of exceeding one of the specific advisory microcystin concentrations of 0.3, 1, 1.6, and 2 g/L is associatied with chlorophyll *a* concentration thresholds of 24.64, 65.6, 89.71, and 94.93, respectively. When managing for these various microcystin levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action."
---

>>>>

---
title: Associations between chlorophyll *a* and various microcystin-LR health advisory
  concentrations
author:
- affilnum: 1
  email: hollister.jeff@epa.gov
  name: Jeffrey W. Hollister
- affilnum: 1
  name: Betty J. Kreakie
affiliation:
- affil: "US Environmental Protection Agency, Office of Research and Development, National
    Health and Environmental Effects Research Laboratory, Atlantic Ecology Division,
    27 Tarzwell Drive  Narragansett, RI, 02882, USA"
  affilnum: 1
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: f1000research.latex
  html_document: null
capsize: normalsize
csl: plos.csl
documentclass: article
fontsize: 11pt
linenumbers: yes
bibliography: manuscript.bib
spacing: doublespacing
abstract: "Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins. To help protect against these impacts, several health advisory levels have been set for some toxins. In particular, one of the more common toxins, microcystin-LR, has several advisory levels set for drinking water and recreational use. However, compared to other water quality measures, field measurements of microcystin-LR are not commonly available due to cost and advanced understanding required to interpret results. Addressing these issues will take time and resources. Thus, there is utility in finding indicators of microcystin-LR that are already widely available, can be estimated quickly and in situ, and used as a first defense against high concentrations of microcystin-LR. Chlorophyll *a* is commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin-LR. In this paper, we use this association to provide estimates of chlorophyll *a* concentrations that are indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR. Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin-LR health advisory level. We look at the recent US EPA health advisories for drinking water as well as the World Health Organization levels for drinking water and recreational use and identify a range of chlorophyll *a* thresholds. A 50% chance of exceeding one of the specific advisory microcystin-LR concentrations of 0.3, 1, 1.6, and 2 g/L is associatied with chlorophyll *a* concentration thresholds of 24.64, 65.6, 89.71, and 94.93, respectively. When managing for these various microcystin-LR levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action."
---

### 2016-01-25 manuscript/f1000_submission/manuscript.tex
fixed table labels
updated letter

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and in situ, and used as a
first defense against high levels of microcystin. Chlorophyll \emph{a}
is commonly measured, can be estimated \emph{in situ}, and has been
shown to be positively associated with microcystin. In this paper, we
use this association to provide estimates of chlorophyll a
concentrations that are indicative of a higher probability of exceeding
select health advisory concentrations for microcystin-LR. Using the 2007
National Lakes Assessment and a conditional probability approach, we
identify chlorophyll \emph{a} concentrations that are more likely than
not to be associated with an exceedance of a microcystin health advisory
level. We look at the recent US EPA health advisories for drinking water
as well as the World Health Organization levels for drinking water and
recreational use and identify a range of chlorophyll \emph{a}
thresholds. A 50\% chance of exceeding one of the specific advisory
microcystin concentrations of 0.3, 1, 1.6, and 2 g/L is associatied with
chlorophyll \emph{a} concentration thresholds of 24.64, 65.6, 89.71, and
94.93, respectively. When managing for these various microcystin levels,
exceeding these reported chlorophyll \emph{a} concentrations should be a
trigger for further testing and possible management action.

@@

\begin{table}[ht]
\caption{Various suggested microcystin-LR health advisory 
                     concentrations from the USEPA and World Health Organization.} 
\label{microcystin_levels}
\begin{tabular}{lll}
  \hline
Source & Type & Concentration \\ 
  \hline
USEPA & Adult Drinking Water Advisory & 1.6 $\mu$g/L \\ 
  USEPA & Child Drinking Water Advisory & 0.3 $\mu$g/L \\ 
  WHO & Drinking Water & 1 $\mu$g/L \\ 
  WHO & Recreational: High Prob. of Effect & 20-2000 $\mu$g/L \\ 
  WHO & Recreational: Low Prob. of Effect & 2-4 $\mu$g/L \\ 
  WHO & Recreational: Moderate Prob. of Effect & 10-20 $\mu$g/L \\ 
  WHO & Recreational: Very High Prob. of Effect & >2000 $\mu$g/L \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Chlorophyll \textit{a} concentrations that are 
                     associated with a 50% probability of exceeding a 
                     microcystin-LR health advisory concentration.} 
\label{mc_chla_table}
\begin{tabular}{rrrrr}
  \hline
Cond. Probability & USEPA Child & WHO Drink & USEPA Adult & WHO Recreational \\ 
  \hline
0.10 & 0.07 & 0.07 & 0.07 & 1.22 \\ 
  0.20 & 0.07 & 4.43 & 11.80 & 20.00 \\ 
  0.30 & 2.98 & 20.43 & 31.40 & 53.64 \\ 
  0.40 & 10.67 & 42.24 & 66.96 & 82.22 \\ 
  0.50 & 23.36 & 66.96 & 83.52 & 105.84 \\ 
  0.60 & 38.30 & 103.20 & 133.20 & 155.52 \\ 
  0.70 & 65.60 & 166.63 & 871.20 & 216.00 \\ 
  0.80 & 117.50 & 338.40 & 871.20 & 871.20 \\ 
  0.90 & 167.04 & 516.00 & 871.20 & 871.20 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the USEPA childrens drinking water advisory.} 
\label{child_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 649 &  96 & 745 \\ 
  Exceed & 162 & 121 & 283 \\ 
  Column Totals & 811 & 217 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the WHO drinking water advisory.} 
\label{who_drink_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 839 &  77 & 916 \\ 
  Exceed &  68 &  44 & 112 \\ 
  Column Totals & 907 & 121 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the USEPA adult drinking water advisory.} 
\label{adult_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 883 &  57 & 940 \\ 
  Exceed &  54 &  34 &  88 \\ 
  Column Totals & 937 &  91 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the WHO recreational water advisory.} 
\label{who_rec_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 910 &  52 & 962 \\ 
  Exceed &  43 &  23 &  66 \\ 
  Column Totals & 953 &  75 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin-LR, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin-LR are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin-LR that are
already widely available, can be estimated quickly and in situ, and used
as a first defense against high concentrations of microcystin-LR.
Chlorophyll \emph{a} is commonly measured, can be estimated \emph{in
situ}, and has been shown to be positively associated with
microcystin-LR. In this paper, we use this association to provide
estimates of chlorophyll \emph{a} concentrations that are indicative of
a higher probability of exceeding select health advisory concentrations
for microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach, we identify chlorophyll \emph{a}
concentrations that are more likely than not to be associated with an
exceedance of a microcystin-LR health advisory level. We look at the
recent US EPA health advisories for drinking water as well as the World
Health Organization levels for drinking water and recreational use and
identify a range of chlorophyll \emph{a} thresholds. A 50\% chance of
exceeding one of the specific advisory microcystin-LR concentrations of
0.3, 1, 1.6, and 2 g/L is associatied with chlorophyll \emph{a}
concentration thresholds of 24.64, 65.6, 89.71, and 94.93, respectively.
When managing for these various microcystin-LR levels, exceeding these
reported chlorophyll \emph{a} concentrations should be a trigger for
further testing and possible management action.

@@

\begin{table}[ht]
\caption{Various suggested microcystin-LR health advisory 
                     concentrations from the USEPA and World Health Organization.} 
\label{tab:microcystin_levels}
\begin{tabular}{lll}
  \hline
Source & Type & Concentration \\ 
  \hline
USEPA & Adult Drinking Water Advisory & 1.6 $\mu$g/L \\ 
  USEPA & Child Drinking Water Advisory & 0.3 $\mu$g/L \\ 
  WHO & Drinking Water & 1 $\mu$g/L \\ 
  WHO & Recreational: High Prob. of Effect & 20-2000 $\mu$g/L \\ 
  WHO & Recreational: Low Prob. of Effect & 2-4 $\mu$g/L \\ 
  WHO & Recreational: Moderate Prob. of Effect & 10-20 $\mu$g/L \\ 
  WHO & Recreational: Very High Prob. of Effect & >2000 $\mu$g/L \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Chlorophyll \textit{a} concentrations that are 
                     associated with a 50% probability of exceeding a 
                     microcystin-LR health advisory concentration.} 
\label{tab:mc_chla_table}
\begin{tabular}{rrrrr}
  \hline
Cond. Probability & USEPA Child & WHO Drink & USEPA Adult & WHO Recreational \\ 
  \hline
0.10 & 0.07 & 0.07 & 0.07 & 1.22 \\ 
  0.20 & 0.07 & 4.43 & 11.80 & 20.00 \\ 
  0.30 & 2.98 & 20.43 & 31.40 & 53.64 \\ 
  0.40 & 10.67 & 42.24 & 66.96 & 82.22 \\ 
  0.50 & 23.36 & 66.96 & 83.52 & 105.84 \\ 
  0.60 & 38.30 & 103.20 & 133.20 & 155.52 \\ 
  0.70 & 65.60 & 166.63 & 871.20 & 216.00 \\ 
  0.80 & 117.50 & 338.40 & 871.20 & 871.20 \\ 
  0.90 & 167.04 & 516.00 & 871.20 & 871.20 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the USEPA childrens drinking water advisory.} 
\label{tab:child_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 649 &  96 & 745 \\ 
  Exceed & 162 & 121 & 283 \\ 
  Column Totals & 811 & 217 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the WHO drinking water advisory.} 
\label{tab:who_drink_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 839 &  77 & 916 \\ 
  Exceed &  68 &  44 & 112 \\ 
  Column Totals & 907 & 121 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the USEPA adult drinking water advisory.} 
\label{tab:adult_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 883 &  57 & 940 \\ 
  Exceed &  54 &  34 &  88 \\ 
  Column Totals & 937 &  91 & 1028 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Confusion matrix comparing chlorophyll \textit{a} 
                     predicted exceedences (rows) versus real exceedances 
                     (columns) for the WHO recreational water advisory.} 
\label{tab:who_rec_conmat_table}
\begin{tabular}{rrrr}
  \hline
 & Not Exceed & Exceed & Row Totals \\ 
  \hline
Not Exceed & 910 &  52 & 962 \\ 
  Exceed &  43 &  23 &  66 \\ 
  Column Totals & 953 &  75 & 1028 \\ 
   \hline
\end{tabular}
\end{table}
